---
title: "Regression Model Course - Quiz1"
author: "Shiv"
date: "Friday, July 31, 2015"
output: html_document
---
## Motivation
Exploration pad for assignments for Quiz 1 of Regressin Model Course.


**Question 1 : We know that minimizer for least squares ($\sum_{i=1}^n (X_i - \mu)^2$) is $\mu$ = $\bar X$. What's the value of $\mu$ if each of the value of $X_i$ have associated weights $W_i = W_1, W_2, \ldots, W_n$**
So we have to find the value of $\mu$ to minimize $\sum_{i=1}^nW_i(X_i-\mu)^2$


Each $X_i$ that has weight$W_i$, can be considered as $W_i$ individual $X_i$'s.
So with this we expand the polulation from N to $\sum_{i=1}^nW_i$. Now the problem is reduced to simple least squares with additional elements, for which $\mu$ is mean. In this case mean would translate to the weighted mean $(\sum_{i=1}^n W_i*X_i)/\sum_{i=1}^nW_i$

``` {r}
x <- c(0.18, -1.54, 0.42, 0.95)
w <- c(2, 1, 3, 1)

weighted.mean(x,w)
```

**Question 2: How to find a regression line that passes thru the origin?**

R prvoides a "lm" function that can be made to not have an intecept and just calculate
a regression line thru the origin.

```{r}
x <- c(0.8, 0.47, 0.51, 0.73, 0.36, 0.58, 0.57, 0.85, 0.44, 0.42)
y <- c(1.39, 0.72, 1.55, 0.48, 1.19, -1.59, 1.23, -0.65, 1.49, 0.05)
lm(y~x-1)

```

**Question 3: Find the slope and intercepts for the regression line. 
User MtCars data set, MPG as outcome and horsepower (hp) as predictor.**

R prvoides a "lm" function that can be used to get the intercept and slope parameters
for the regression line.

```{r}
lm(mtcars$mpg~mtcars$hp)
```
Also, the regression line may be defined as $Y_i=\beta_0 + \beta_1*X_i$ where $\beta_0$
is the intercept and $\beta_1$ is the slope. 
$$
\beta_1 = Cor(y,x)*std(y)/std(x)    \beta_0 = \bar Y - \beta_1*\bar X
$$

```{r}
slope<-cor(mtcars$mpg,mtcars$hp)*sd(mtcars$mpg)/sd(mtcars$hp)
slope
intercept<-mean(mtcars$mpg) - mean(mtcars$hp)*slope
intercept
```

**Question 6: What is Normalization? How to calcualte a normalized value for 
a given variable**

"Normalization is process of adjusting values measured of diffirent scales to a notinonally common scale"
[wikipedia] (https://en.wikipedia.org/wiki/Normalization_(statistics))

This can be achived in R using the *scale* function
```{r}
x <- c(8.58, 10.46, 9.01, 9.64, 8.86)
str(scale(x))
```
Normalized value $Z_i = (X_i - \bar X)/std(X)$
``` {r}
x <- c(8.58, 10.46, 9.01, 9.64, 8.86)
round((8.58 - mean(x))/sd(x),3)
````